SNP phylogeny

### 2023-05-20 episodes/data-visualization.Rmd
Draft of data viz
Added conclusion, moved files around, first complete draft

---
title: 'episode_example'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions

- How do you write a lesson using R Markdown and `{sandpaper}`?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Explain how to use markdown with the new lesson template
- Demonstrate how to include pieces of code, figures, and nested challenge blocks

::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

This is a lesson created via The Carpentries Workbench. It is written in
[Pandoc-flavored Markdown][pandoc] for static files (with extension `.md`) and
[R Markdown][r-markdown] for dynamic files that can render code into output
(with extension `.Rmd`). Please refer to the [Introduction to The Carpentries
Workbench][carpentries-workbench] for full documentation.

What you need to know is that there are three sections required for a valid
Carpentries lesson template:

 1. `questions` are displayed at the beginning of the episode to prime the
    learner for the content.
 2. `objectives` are the learning objectives for an episode displayed with
    the questions.
 3. `keypoints` are displayed at the end of the episode to reinforce the
    objectives.

:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor

Inline instructor notes can help inform instructors of timing challenges
associated with the lessons. They appear in the "Instructor View"

::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 1: Can you do it?

What is the output of this command?

```r
paste("This", "new", "lesson", "looks", "good")
```

:::::::::::::::::::::::: solution

## Output

```output
[1] "This new lesson looks good"
```

:::::::::::::::::::::::::::::::::

## Challenge 2: how do you nest solutions within challenge blocks?

:::::::::::::::::::::::: solution

You can add a line with at least three colons and a `solution` tag.

:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

## Figures

You can include figures generated from R Markdown:

```{r pyramid, fig.alt = "pie chart illusion of a pyramid", fig.cap = "Sun arise each and every morning"}
pie(
  c(Sky = 78, "Sunny side of pyramid" = 17, "Shady side of pyramid" = 5), 
  init.angle = 315, 
  col = c("deepskyblue", "yellow", "yellow3"), 
  border = FALSE
)
```
Or you can use pandoc markdown for static figures with the following syntax:

`![optional caption that appears below the figure](figure url){alt='alt text for
accessibility purposes'}`

![You belong in The Carpentries!](https://raw.githubusercontent.com/carpentries/logo/master/Badge_Carpentries.svg){alt='Blue Carpentries hex person logo with no text.'}

## Math

One of our episodes contains $\LaTeX$ equations when describing how to create
dynamic reports with {knitr}, so we now use mathjax to describe this:

`$\alpha = \dfrac{1}{(1 - \beta)^2}$` becomes: $\alpha = \dfrac{1}{(1 - \beta)^2}$

Cool, right?

::::::::::::::::::::::::::::::::::::: keypoints

- Use `.md` files for episodes when you want static content
- Use `.Rmd` files for episodes when you need to generate output
- Run `sandpaper::check_lesson()` to identify any issues with your lesson
- Run `sandpaper::build_lesson()` to preview your lesson locally

::::::::::::::::::::::::::::::::::::::::::::::::

>>>>

---
title: 'Data Visualization'
teaching: 10
exercises: 60
---

## Comparison of clustering results

::::::::::::::::::::::::::::::::::::::: objectives

- Explore the resulting trees in conjunction with the meta data.
- Make an estimation on the likelihood of transmission events

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- How are phylogenetic trees viewed and compared?
- How can I visualize several layers of data?

::::::::::::::::::::::::::::::::::::::::::::::::::

We grouped (clustered) our isolates by the information that we extracted from the sequencing reads. We compared our isolates to a reference genome and clustered them on basis of the single nucleotide variations that they exhibit when compared to this reference.

![](fig/Workflow.png){alt='workflow'}

Next, we will visualize this clustering in the context of the meta data. Make sure you have the SNP tree that you generated (or can be downloaded here [core.aln.newick](files/core.aln.newick)) on your own computer. Download also the [metadata file](files/Molepi_meta.csv).

## Visualization of genetic information and metadata

Visualization is frequently used to aid the interpretation of complex datasets. Within microbial genomics, visualizing the relationships between multiple genomes as a tree provides a framework onto which associated data (geographical, temporal, phenotypic and epidemiological) are added to generate hypotheses and to explore the dynamics of the system under investigation.

1. Open Chrome. Go to [microreact](https://microreact.org/). 
2. Click on 'Upload'. Drag-and-drop the newick tree produced by snippy and IQ-TREE ([core.aln.newick](files/core.aln.newick)). 
3. Click on "Add more files" and select the the [metadata file](files/Molepi_meta.csv), which is a table comprising metadata file. Choose 'Data (csv or tsv)' as file kind.
4. Click on continue.

First, the tree should be rooted with the reference strain (the one that has the longest branch). Hover over the middle of the long branch, until a circle appears. Right-click on it, and select "Set as root (re-root)".

Explore the location, time and further meta data. Play around with the different visualization options with the help of the [documentation of microreact](https://docs.microreact.org/).

::::::::::::::::::::::::::::::::::::::  discussion

## Discussion

1. Which transmission events are likely based on the metadata (location, RFLP, date) alone?
2. Which transmission events are likely based on the SNP data?
3. Draw a transmission tree if possible.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: keypoints

- Genetic information can confirm or contradict the meta data

::::::::::::::::::::::::::::::::::::::::::::::::::

### 2023-05-20 episodes/snps-phylogeny.Rmd
Added conclusion, moved files around, first complete draft

Phylogenetic trees have been discussed during the lectures. We will here establish a phylogenetic tree from the file 'core.aln' with [IQ-TREE](http://www.iqtree.org/). IQ-TREE is a phylogeny software based on the maximum-likelihood principle. IQ-TREE has been widely used and cited. There is a very wide range of parameters that need to be chosen, such as nucleotide or amino-acid substitution models.

@@

Let's inspect our tree.

```bash
$ cd ~/molepi/results/
$ cd tree
$ cat core.aln.treefile
```

@@

```bash
$ cat core.aln.treefile
```

@@

IQ-TREE chose the TVMe+ASC model. We've discussed the ASC above, and you can read more about the TVMe in the [IQ-TREE manual](http://www.iqtree.org/doc/Substitution-Models). In short, it is a mdoel where base frequencies are deemed equal, and all rates estimated independently, except the rates of transversions (A <-> G and C <-> T) are equal.

>>>>

Phylogenetic trees have been discussed during the lectures. We will here establish a phylogenetic tree from the file 'core.aln' with [IQ-TREE](https://www.iqtree.org/). IQ-TREE is a phylogeny software based on the maximum-likelihood principle. IQ-TREE has been widely used and cited. There is a very wide range of parameters that need to be chosen, such as nucleotide or amino-acid substitution models.

@@

Let's inspect our tree, and give it another extension to make it clear it is a newick file.

```bash
$ cd ~/molepi/results/
$ cd tree
$ mv core.aln.treefile core.aln.newick
$ core.aln.newick
```

@@

```bash
$ cat core.aln.iqtree
```

@@

IQ-TREE chose the TVMe+ASC model. We've discussed the ASC above, and you can read more about the TVMe in the [IQ-TREE manual](https://www.iqtree.org/doc/Substitution-Models). In short, it is a mdoel where base frequencies are deemed equal, and all rates estimated independently, except the rates of transversions (A <-> G and C <-> T) are equal.
